Ice-shelf ocean boundary layer dynamics from large-eddy simulations

### 2021-10-19 SI_rev1.tex
Revisions for review round 1

@@

\begin{figure}[t]
\includegraphics[width=6cm]{figS7.pdf}
\caption{Relationship between far-field thermal driving and melt rate. This figure is the same as Figure 7a but values are averaged over each inertial cycle. The largest points correspond to the fourth and last inertial cycle with progressively smaller points for previous inertial cycles. }
\label{fig:melt_sensitivity_cycles}
\end{figure}

>>>>

\begin{figure}[t]
\includegraphics[width=12cm]{figS8.pdf}
\caption{Sensitivity of simulated mean state and turbulent fluxes to resolution. The vertical resolution is double the horizontal resolution ($\Delta x, \Delta y$). Results are averaged over the first inertial period after a 2\unit{h} spin-up. (a) Temperature. (b) Salinity. (c) Total vertical heat flux (solid), resolved vertical heat flux (dashed), and sub-grid vertical heat flux (dotted). (d) Velocity, u-component (solid) and v-component (dashed).}
\label{fig:res_sensitivity}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=6cm]{figS7.pdf}
\caption{Relationship between far-field thermal driving and melt rate. This figure is the same as Figure 7a but values are averaged over each inertial cycle. The largest points correspond to the fourth and last inertial cycle with progressively smaller points for previous inertial cycles.}
\label{fig:melt_sensitivity_cycles}
\end{figure}

### 2021-10-19 main_rev1.tex
Revisions for review round 1

The turbulence closure scheme is Anisotropic Minimum Dissipation \cite[AMD;][]{rozema_minimum-dissipation_2015} employing the extension to scalars introduced by \citet{abkar_minimum-dissipation_2016}. We validated our implementation against the stable atmospheric boundary layer test case published in \citet{abkar_large-eddy_2017} and \citet{stoll_large-eddy_2008}. We follow \citet[]['s]{abkar_large-eddy_2017}'s model set-up exactly and chose their moderate resolution with 72 x 72 x 72 grid points. Our results are in good agreement with \citet[]['s]{abkar_large-eddy_2017} solution employing another LES model with AMD (boundary layer height differs by <10\%) and other LES models with different turbulence closures (Fig. S1, their Fig. 1 and references therein). Typical sub-grid scale momentum and scalar diffusivities in our sub-ice-shelf simulations of stratified turbulence range from $10^{-5}$ to $10^{-4}$\,\unit{m^2\,s^{-1}} in the upper two-thirds of the domain where damping is not applied (the damping methodology is discussed in Section \ref{meth:setup}).

@@

\subsection{Overview of the mean simulated state}
% Description of time evolution
Melt rates decline over the course of the simulations (Figure~\ref{fig:timeseries}), preventing the identification of steady-state melt rates for most runs. This decline is in response to the concomitant increase in stratification during the course of the simulations, which decreases vertical heat fluxes by reducing vertical turbulent fluctuations. We do not continue our simulations beyond 4 inertial periods in the hope of reaching steady-state conditions because the increase in stratification reduces the turbulent length scales and necessitates higher model resolution than we could computationally afford. Our simulations generally show resolved turbulent fluxes exceeding subgrid turbulent fluxes by at least a factor of two, but subgrid turbulent fluxes dominate within several meters of the ice base where the stratification is strongest (Figure~S1). Resolved turbulent fluxes are also comparable to subgrid turbulent fluxes late in the simulation of low slope cases after significant TKE has been lost ($\leq\!0.1$\unit{^{\circ}C} slopes, Figure~S1d).

\begin{figure*}[t]
\includegraphics[width=12cm]{fig1_rev1.pdf}
\caption{Time evolution of (a,d) domain-averaged, resolved turbulence kinetic energy (TKE), (b,e) melt rate and (c,f) friction velocity for (a-c) thermal driving simulations and (d-f) variable slope simulations. The black curve represents the same simulation in all panels in this and subsequent figures. The analysis window, the last inertial period, is shaded.}
\label{fig:timeseries}
\end{figure*}

@@

Here, we characterize the evolution of resolved TKE and primes designate resolved fluctuations. Figure \ref{fig:tke_budget}b-d shows the source terms in this budget for the variable slope simulations; the analogous figure for the thermal driving simulations is Figure~S2 and shows similar patterns. Shear production ($F_{shear}$) ranges from $\sim 10^{-9} \textrm{--} 10^{-8}$\,\unit{m^2\,s^{-3}} with a maximum at the ice-ocean interface and local maxima in the upslope flow within 5\,\unit{m} of the boundary (Figure \ref{fig:tke_budget}a). The increase in TKE throughout the boundary layer as slope increases reflects this shear-induced turbulence (Figure \ref{fig:tke_budget}a).

@@

\subsection{Boundary layer turbulence}
% discuss turbulent structures
To demonstrate the simulated turbulent structures in this regime, we present horizontal and vertical cross-sectional  snapshots through the domain for the two slope end-members in Figure \ref{fig:cross-section}. Turbulent structures within the IOBL are consistent with propagating Holmboe shear instabilities under stable stratification \citep{carpenter_identifying_2010}. Shear is stronger within the IOBL than at the base of the IOBL due to the concentration of buoyant plume flow near the top of the IOBL. Consequently, the amplitude of these structures increases near the boundary for the more strongly-sloped simulations (Figures \ref{fig:cross-section}). The difference in IOBL turbulence with slope is perhaps best seen in the turbulent structures at 1\,\unit{m} from the ice-ocean interface (Figure \ref{fig:cross-section}e,f). The structures become increasingly filamentous (i.e., near-wall streaks, e.g., \citet{del_alamo_spectra_2003, hoyas_scaling_2006}) and coherent as slope and thus the velocity of the buoyancy-driven current increase. Since the stratification decreases with increasing slope, the ratio of vertical to horizontal velocity variance also increases; velocity fluctuations are less confined to the slope-parallel direction (Figure~S3).

@@

\begin{figure*}[t]
\includegraphics[width=12cm]{fig6.pdf}
\caption{Horizontally-averaged turbulence characteristics for (a) 0.1\unit{^{\circ}C} thermal driving, 0.01\unit{^{\circ}} slope case and (b) 0.6\unit{^{\circ}C} thermal driving, 1\unit{^{\circ}} slope case. Turbulence is considered intermittent when the dissipation contour of $10^{-9}$\,\unit{m^2\,s^{-3}} (dashed line) reaches the boundary. Significant turbulence kinetic energy (TKE, green-yellow shading) can be present when dissipation is low. Higher TKE at -25\,\unit{m} at later times in (a) is due to turbulent transport, while the higher TKE at later times in (b) is due to shear production. IOBL depth is contoured in black. }
\label{fig:intermittency}
\end{figure*}

@@

Vertical heat fluxes are shown in Figure \ref{fig:flux_profiles}b,e. The vertical heat flux has a maximum at the ice-ocean interface and decreases throughout the IOBL, with small values below the pycnocline. Since the IOBL is fully turbulent (with the exception of some intermittency discussed below), the sub-grid diffusivities of momentum, heat and salt closely resemble one another (Figure~S4). The vertical salt flux profiles are shown in Figure~S5 and have a very similar shape to the vertical heat flux profiles.

% Scaling for flux profiles
The relatively narrow range of conditions simulated here suggests that a depth-dependent shape function for scalar fluxes could be formulated. The distance from the interface is scaled by the Ekman depth:
\begin{equation}
    d_E = (2 K_e)^{1/2} |f_3|^{-1/2}
\end{equation}
where $K_e$ is the mean eddy viscosity in the turbulent boundary layer assuming the total fluxes follow Fick's law:
\begin{equation}
    \overline{w'\vec{u}'} = K_e d\vec{u}/dz
\end{equation}
$K_e$ profiles are shown in Figure~S6.

@@

% Argument for validity of findings in non-steady conditions
The decline in IOBL turbulence discussed in Section \ref{disc:dyn} results in declining melt rates. Thus, we could not evaluate the relationship between melting and far-field conditions at steady state, which would have offered the most direct path to assessing melt rate sensitivity to ocean conditions. As discussed in \citet{jenkins_simple_2016}, achieving steady-state solutions in simulations may require prescribing large-scale gradients in temperature. We have not included these large-scale gradients in our simulations, but this may be an avenue for future work. However, we believe that our transient solutions do provide some indications of how melt rates and boundary layer properties depend on ocean temperature and ice-shelf slopes. One justification for this belief is that the simulated sensitivity of melt rate to ocean temperature remains consistently linear through all inertial periods simulated (Figure~S7). On the other hand, the differently sloped simulations continue to diverge at the end of the simulation as the IOBL accelerates (Figure \ref{fig:timeseries}d-f). Nonetheless we are able to find relationships between the vertical heat flux profiles and ocean temperature and slope that suggest predictability of the mean effects of turbulence despite the transience of these simulations.

@@

% thermal exchange coeff, slope
We find a linear relationship between the derived thermal exchange coefficient and the slope of the ice-shelf base, indicating that mixing is more efficient at higher slopes. This relationship also holds for the derived friction velocity from a quadratic drag law (Figure \ref{fig:melt_sensitivity}d). Note that these quadratic-drag friction velocities are greater than the model's parameterization of friction velocity as the former neglects stratification effects while the latter includes them. This enhanced turbulent efficiency is due to enhanced shear instabilities, and is reflected in turbulent diffusivities parameterized by the AMD scheme. Much greater variability in sub-grid turbulent diffusivities is seen over the variation in slope than the variation in thermal driving tested here (Figure~S4).

>>>>

The turbulence closure scheme is Anisotropic Minimum Dissipation \cite[AMD;][]{rozema_minimum-dissipation_2015} employing the extension to scalars introduced by \citet{abkar_minimum-dissipation_2016}. We validated our implementation against the stable atmospheric boundary layer test case published in \citet{abkar_large-eddy_2017} and \citet{stoll_large-eddy_2008}. We follow \citet[]['s]{abkar_large-eddy_2017}'s model set-up exactly and chose their moderate resolution with 72 x 72 x 72 grid points. Our results are in good agreement with \citet[]['s]{abkar_large-eddy_2017} solution employing another LES model with AMD (boundary layer height differs by <10\%) and other LES models with different turbulence closures (Figure~S1, their Fig. 1 and references therein). Typical sub-grid scale momentum and scalar diffusivities in our sub-ice-shelf simulations of stratified turbulence range from $10^{-5}$ to $10^{-4}$\,\unit{m^2\,s^{-1}} in the upper two-thirds of the domain where damping is not applied (the damping methodology is discussed in Section \ref{meth:setup}).

@@

\subsection{Overview of the mean simulated state}
% Description of time evolution
Melt rates decline over the course of the simulations (Figure~\ref{fig:timeseries}), preventing the identification of steady-state melt rates for most runs. This decline is in response to the concomitant increase in stratification during the course of the simulations, which decreases vertical heat fluxes by reducing vertical turbulent fluctuations. We do not continue our simulations beyond 4 inertial periods in the hope of reaching steady-state conditions because the increase in stratification reduces the turbulent length scales and necessitates higher model resolution than we could computationally afford. Our simulations generally show resolved turbulent fluxes exceeding subgrid turbulent fluxes by at least a factor of two, but subgrid turbulent fluxes dominate within several meters of the ice base where the stratification is strongest (Figure~S2). Resolved turbulent fluxes are also comparable to subgrid turbulent fluxes late in the simulation of low slope cases after significant TKE has been lost ($\leq\!0.1$\unit{^{\circ}C} slopes, Figure~S2d). We evaluated the effects of doubling both horizontal and vertical resolution in a separate simulation with an otherwise identical set-up to the base case, run for one inertial period after spin-up. While a greater portion of the vertical heat fluxes was resolved as expected, the melt rate averaged over one inertial period only increased by 7\% and differences in the mean state were sufficiently small to justify the use of our standard resolution (Figure~S3).

\begin{figure*}[t]
\includegraphics[width=12cm]{fig1.pdf}
\caption{Time evolution of (a,d) domain-averaged, resolved turbulence kinetic energy (TKE), (b,e) melt rate and (c,f) friction velocity for (a-c) thermal driving simulations and (d-f) variable slope simulations. The black curve represents the same simulation in all panels in this and subsequent figures. The analysis window, the last inertial period, is shaded.}
\label{fig:timeseries}
\end{figure*}

@@

Here, we characterize the evolution of resolved TKE and primes designate resolved fluctuations. Figure \ref{fig:tke_budget}b-d shows the source terms in this budget for the variable slope simulations; the analogous figure for the thermal driving simulations is Figure~S4 and shows similar patterns. Shear production ($F_{shear}$) ranges from $\sim 10^{-9} \textrm{--} 10^{-8}$\,\unit{m^2\,s^{-3}} with a maximum at the ice-ocean interface and local maxima in the upslope flow within 5\,\unit{m} of the boundary (Figure \ref{fig:tke_budget}a). The increase in TKE throughout the boundary layer as slope increases reflects this shear-induced turbulence (Figure \ref{fig:tke_budget}a).

@@

\subsection{Boundary layer turbulence}
% discuss turbulent structures
To demonstrate the simulated turbulent structures in this regime, we present horizontal and vertical cross-sectional  snapshots through the domain for the two slope end-members in Figure \ref{fig:cross-section}. Turbulent structures within the IOBL are consistent with propagating Holmboe shear instabilities under stable stratification \citep{carpenter_identifying_2010}. Shear is stronger within the IOBL than at the base of the IOBL due to the concentration of buoyant plume flow near the top of the IOBL. Consequently, the amplitude of these structures increases near the boundary for the more strongly-sloped simulations (Figures \ref{fig:cross-section}). The difference in IOBL turbulence with slope is perhaps best seen in the turbulent structures at 1\,\unit{m} from the ice-ocean interface (Figure \ref{fig:cross-section}e,f). The structures become increasingly filamentous (i.e., near-wall streaks, e.g., \citet{del_alamo_spectra_2003, hoyas_scaling_2006}) and coherent as slope and thus the velocity of the buoyancy-driven current increase. Since the stratification decreases with increasing slope, the ratio of vertical to horizontal velocity variance also increases; velocity fluctuations are less confined to the slope-parallel direction (Figure~S5).

@@

\begin{figure*}[t]
\includegraphics[width=12cm]{fig6.pdf}
\caption{Horizontally-averaged turbulence characteristics for (a) 0.15\unit{^{\circ}C} thermal driving, 0.01\unit{^{\circ}} slope case and (b) 0.6\unit{^{\circ}C} thermal driving, 1\unit{^{\circ}} slope case. Turbulence is considered intermittent when the dissipation contour of $10^{-9}$\,\unit{m^2\,s^{-3}} (dashed line) reaches the boundary. Significant turbulence kinetic energy (TKE, green-yellow shading) can be present when dissipation is low. Higher TKE at -25\,\unit{m} at later times in (a) is due to turbulent transport, while the higher TKE at later times in (b) is due to shear production. IOBL depth is contoured in black. }
\label{fig:intermittency}
\end{figure*}

@@

Vertical heat fluxes are shown in Figure \ref{fig:flux_profiles}b,e. The vertical heat flux has a maximum at the ice-ocean interface and decreases throughout the IOBL, with small values below the pycnocline. Since the IOBL is fully turbulent (with the exception of some intermittency discussed below), the sub-grid diffusivities of momentum, heat and salt closely resemble one another (Figure~S6). The vertical salt flux profiles are shown in Figure~S7 and have a very similar shape to the vertical heat flux profiles.

% Scaling for flux profiles
The relatively narrow range of conditions simulated here suggests that a depth-dependent shape function for scalar fluxes could be formulated. The distance from the interface is scaled by the Ekman depth:
\begin{equation}
    d_E = (2 K_e)^{1/2} |f_3|^{-1/2}
\end{equation}
where $K_e$ is the mean eddy viscosity in the turbulent boundary layer assuming the total fluxes follow Fick's law:
\begin{equation}
    \overline{w'\vec{u}'} = K_e d\vec{u}/dz
\end{equation}
$K_e$ profiles are shown in Figure~S8.

@@

% Argument for validity of findings in non-steady conditions
The decline in IOBL turbulence discussed in Section \ref{disc:dyn} results in declining melt rates. Thus, we could not evaluate the relationship between melting and far-field conditions at steady state, which would have offered the most direct path to assessing melt rate sensitivity to ocean conditions. As discussed in \citet{jenkins_simple_2016}, achieving steady-state solutions in simulations may require prescribing large-scale gradients in temperature. We have not included these large-scale gradients in our simulations, but this may be an avenue for future work. However, we believe that our transient solutions do provide some indications of how melt rates and boundary layer properties depend on ocean temperature and ice-shelf slopes. One justification for this belief is that the simulated sensitivity of melt rate to ocean temperature remains consistently linear through all inertial periods simulated (Figure~S9). On the other hand, the differently sloped simulations continue to diverge at the end of the simulation as the IOBL accelerates (Figure \ref{fig:timeseries}d-f). Nonetheless we are able to find relationships between the vertical heat flux profiles and ocean temperature and slope that suggest predictability of the mean effects of turbulence despite the transience of these simulations.

@@

% thermal exchange coeff, slope
We find a linear relationship between the derived thermal exchange coefficient and the slope of the ice-shelf base, indicating that mixing is more efficient at higher slopes. This relationship also holds for the derived friction velocity from a quadratic drag law (Figure \ref{fig:melt_sensitivity}d). Note that these quadratic-drag friction velocities are greater than the model's parameterization of friction velocity as the former neglects stratification effects while the latter includes them. This enhanced turbulent efficiency is due to enhanced shear instabilities, and is reflected in turbulent diffusivities parameterized by the AMD scheme. Much greater variability in sub-grid turbulent diffusivities is seen over the variation in slope than the variation in thermal driving tested here (Figure~S6).
